---
title: "COVID-19 Data Project"
author: "Ryan Scott"
date: "2025-06-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

### Statement of Interest

The following project is an Exploratory Data Analysis of COVID-19.

### Background of Source Data 

The data set for this project was obtained from the John Hopkins website.


We will be using the tidyverse library for this project.

```{r tidyverse}
#install.packages("tidyverse")
library(tidyverse)
```

I assign each csv file to a variable.

```{r variables}
global_cases <- read_csv("~/Desktop/Documents/Data Analysis Tools/Projects/RStudio Projects/COVID/time_series_covid19_confirmed_global.csv")
global_deaths <- read_csv("~/Desktop/Documents/Data Analysis Tools/Projects/RStudio Projects/COVID/time_series_covid19_deaths_global.csv")
US_cases <- read_csv("~/Desktop/Documents/Data Analysis Tools/Projects/RStudio Projects/COVID/time_series_covid19_confirmed_US.csv")
US_deaths <- read_csv("~/Desktop/Documents/Data Analysis Tools/Projects/RStudio Projects/COVID/time_series_covid19_deaths_US.csv")

```

I first want to look the columns for global cases and global deaths. I use the head function to see the column names with a set number of rows.

```{r}
head(global_cases, n = 5)
```

```{r}
head(global_deaths, n=5)
```

This code reshapes the global cases data set by converting daily case columns into a longer, tidy format where each row represents a specific location and date. It removes latitude and longitude columns, streamlining the data for easier analysis or visualization.

```{r}
global_cases <- global_cases %>%
     pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
                  names_to = "date",
                 values_to = "cases") %>%
     select(-c(Lat, Long))
```

This code reshapes the global deaths data set by converting daily case columns into a longer, tidy format where each row represents a specific location and date. It removes latitude and longitude columns, streamlining the data for easier analysis or visualization.

```{r}
global_deaths <- global_deaths %>%
     pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
                  names_to = "date",
                  values_to = "deaths") %>%
     select(-c(Lat, Long))
```


The following code performs a full join to keep all records from both data tables, renames the country and province columns for consistency, and converts the date column from character format to a proper date format using mdy() (month-day-year).

```{r}
global <- global_cases %>%
     full_join(global_deaths) %>%
     rename(Country_Region = 'Country/Region',
            Province_State = 'Province/State') %>%
     mutate(date = mdy(date))
```

We use summary() to give us a quick statistical overview of the global data table

```{r}
summary(global)
```

We filter the global table to give us countries where cases are greater than 0.

```{r}
global <- global %>% filter(cases > 0)
```

We use the head function to view the columns in the US cases data set.

```{r}
head(US_cases, n =5)
```

This code reshapes the US cases data set by converting daily case columns into a longer, tidy format where each row represents a specific location and date. It removes latitude and longitude columns, streamlining the data for easier analysis or visualization.

```{r}
US_cases <- US_cases %>%
    pivot_longer(cols = -(UID:Combined_Key),
                 names_to = "date",
                 values_to = "cases") %>%
    select(Admin2:cases) %>%
    mutate(date = mdy(date)) %>%
    select(-c(Lat, Long_))
```

```{r}
head(US_deaths, n = 5)
```
This code reshapes the US Deaths data set by converting daily case columns into a longer, tidy format where each row represents a specific location and date. It removes latitude and longitude columns, streamlining the data for easier analysis or visualization.

```{r}
US_deaths <- US_deaths %>%
    pivot_longer(cols = -(UID:Population),
                 names_to = "date",
                 values_to = "deaths") %>%
    select(Admin2:deaths) %>%
    mutate(date = mdy(date)) %>%
    select(-c(Lat, Long_))
```

The following code performs a full join to keep all records from both data tables

```{r}
US <- US_cases %>%
     full_join(US_deaths)
```

```{r}
global <- global %>%
     unite("Combined_Key",
          c(Province_State, Country_Region),
          sep = ", ",
          na.rm = TRUE,
          remove = FALSE)
```

```{r}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

```

```{r}
uid <- read_csv(uid_lookup_url) %>%
     select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```

```{r}
global <- global %>%
     left_join(uid, by = c("Province_State", "Country_Region")) %>%
     select(-c(UID, FIPS)) %>%
     select(Province_State, Country_Region, date,
          cases, deaths, Population,
          Combined_Key)
```

```{r}
US_by_state <- US %>%
     group_by(Province_State, Country_Region, date) %>%
     summarize(cases = sum(cases), deaths = sum(deaths), 
               Population = sum(Population))  %>%
     mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
     select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
     ungroup()
```

```{r}
US_totals <- US_by_state %>%
     group_by(Country_Region, date) %>%
     summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
     mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
     select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
     ungroup()
```

```{r}
US_totals %>%
     filter(cases > 0) %>%
     ggplot(aes(x = date, y = cases)) +
     geom_line(aes(color = "cases")) +
     geom_point(aes(color = "cases")) +
     geom_line(aes(y = deaths, color = "deaths")) +
     geom_point(aes(y = deaths, color = "deaths")) +
     scale_y_log10() +
     theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
     labs(title = "COVID19 in US", y = NULL)
```

```{r}
state <- "New York"
US_by_state %>%
     filter(Province_State == state) %>%
     filter(cases > 0) %>%
     ggplot(aes(x = date, y = cases)) + 
     geom_line(aes(color = "cases")) +
     geom_point(aes(color = "cases")) +
     geom_line(aes(y = deaths, color = "deaths")) +
     geom_point(aes(y = deaths, color = "deaths")) +
     scale_y_log10() +
     theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
     labs(title = str_c("COVID19 in ", state), y = NULL)
```

```{r}
max(US_totals$date)
max(US_totals$deaths)
```

Building new columns to determine the number of new cases each day

```{r}
US_by_state <- US_by_state %>%
     mutate(new_cases = cases - lag(cases), 
          new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
     mutate(new_cases = cases - lag(cases),
          new_deaths = deaths - lag(deaths))
```

```{r}
tail(US_totals)
```

```{r}
tail(US_totals %>% select(new_cases, new_deaths, everything()))
```

```{r}
US_totals %>%
     ggplot(aes(x = date, y = new_cases)) +
     geom_line(aes(color = "new_cases")) +
     geom_point(aes(color = "new_cases")) +
     geom_line(aes(y = new_deaths, color = "new_deaths")) +
     geom_point(aes(y = new_deaths, color = "new_deaths")) +
     scale_y_log10() +
     theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
     labs(title = "COVID19 in US", y = NULL)
```

```{r}
state <- "New York"
US_by_state %>%
     filter(Province_State == state) %>%
     ggplot(aes(x = date, y = new_cases)) + 
     geom_line(aes(color = "new_cases")) +
     geom_point(aes(color = "new_cases")) +
     geom_line(aes(y = deaths, color = "new_deaths")) +
     geom_point(aes(y = deaths, color = "new_deaths")) +
     scale_y_log10() +
     theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
     labs(title = str_c("COVID19 in ", state), y = NULL)
```

```{r}
US_state_totals <- US_by_state %>%
     group_by(Province_State) %>%
     summarize(deaths = max(deaths), cases = max(cases),
          population = max(Population),
          cases_per_thou = 1000 * cases / population,
          deaths_per_thou = 1000 * deaths / population) %>%
     filter(cases > 0, population > 0)
```

Tells me the the states with the smallest states of deaths per thousand

```{r}
US_state_totals %>%
     slice_min(deaths_per_thou, n = 10)
```

```{r}
US_state_totals %>%
     slice_min(deaths_per_thou, n = 10) %>%
     select(deaths_per_thou, cases_per_thou, everything())
```

Worst states

```{r}
US_state_totals %>%
     slice_max(deaths_per_thou, n = 10) %>%
     select(deaths_per_thou, cases_per_thou, everything())
```

```{r}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
```

```{r}
US_state_totals %>% slice_min(cases_per_thou)
```

```{r}
US_state_totals %>% slice_max(cases_per_thou)
```

```{r}
US_state_totals %>% mutate(pred = predict(mod))
```

```{r}
us_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
us_tot_w_pred
```

```{r}
us_tot_w_pred %>% 
     ggplot() +
     geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
     geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

